Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-29 Manuscript_Nater_etal.qmd

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but also often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We set up an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis, which predicts higher reproductive output when there is plenty of alternative prey for shared predators. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study thus highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, recent population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a substantial number of areas.

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). For willow ptarmigan, evidence for density-dependent population regulation has been mixed. @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly. While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle. Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A]. When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv"). While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation. Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.

@@

Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small. Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework. Other relevant data could be included using the same approach, for example data from ongoing nesting success monitoring (Nilsen, E.B. unpublished data), data from past studies of marked birds [@sandercock_etal2011], and data from other monitoring programs for breeding birds based on point counts (see e.g. the Norwegian Breeding Bird Monitoring: <https://hekkefuglovervakingen.nina.no/>). The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data. Such data might be available with two different spatial and temporal resolution. First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>). Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters pr area pr day). As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g., @gamelon2021; @nater2021]. While harvest effects on willow ptarmigan have been studied previously, much uncertainty remains [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001]. For example, little is known about how harvest pressure and density feedbacks interact on differente temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018]. Additionally, harvest effects often interact with other (emergent) factors such as climate change and habitat degradation, making predictive models that account for harvest alongside other mechanisms invaluable for informing policy changes [@gamelon2019].

@@

All code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>. The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).

>>>>

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable. Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data. While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.

<!--# removed a few words/phrases that I thought were superfulous-->

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, recent population density appeared to be lowest in the northern Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a substantial number of areas.

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]. For willow ptarmigan, evidence for density-dependent population regulation has been mixed. @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly. While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle. Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A]. When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv"). While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation. Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.

@@

Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small. Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework. Other relevant data could be included using the same approach, for example data from ongoing nesting success monitoring (Nilsen, E.B. unpublished data), data from past studies of marked birds [@sandercock_etal2011], and data from other monitoring programs for breeding birds based on point counts (see e.g. the Norwegian Breeding Bird Monitoring: <https://hekkefuglovervakingen.nina.no/>). The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data. Such data might be available with two different spatial and temporal resolution. First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>). Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters per area per day). As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g., @gamelon2021; @nater2021]. While harvest effects on willow ptarmigan have been studied previously, much uncertainty remains [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001]. For example, little is known about how harvest pressure and density feedbacks interact on different temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018]. Additionally, harvest effects often interact with other (emergent) factors such as climate change and habitat degradation, making predictive models that account for harvest alongside other mechanisms invaluable for informing policy changes [@gamelon2019].

@@

All code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>. The results presented in this paper were created using version 2.0 of the code [@chloernater_2024_10572340].
